### 2010-09-21 doc/guide/language.tex
- added an index covering the language
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/trunk/gringo@3398 3e02f596-3131-0410-8352-e04eeaec37cc

\subsubsection{Normal Programs and Integrity Constraints}\label{subsec:gringo:normal}

@@

\begin{tabular}{rl@{}l}\label{eq:normal:rule}
\textbf{Rule:} & $A_0$& $\code{:-}~L_1\code{,}\dots\code{,}L_n$\code{.}
\\
\textbf{Fact:} & $A_0$&\code{.}
\\
\textbf{Integrity Constraint:} & & $\code{:-}~L_1\code{,}\dots\code{,}L_n$\code{.}
\end{tabular}
\\

\noindent
The head~$A_0$ of a fact or a rule is an \emph{atom} of the same form
as a function symbol or constant.
Any $L_j$ is a \emph{literal} of the form $A$ or $\code{not}~A$
for an atom~$A$ where the connective \code{not} corresponds to default negation.
The set of literals $\{L_1,\dots,L_n\}$ is called the body of the rule.
Facts have an empty body.
Throughout this section we further extend the predicates that can be used 
in a rule including comparison predicates (Section \ref{subsec:gringo:comp}) and aggregates (Section \ref{subsec:gringo:aggregate}).
Furthermore, \gringo{} expects rules to be safe, i.e., 
all variables that appear in a rule have to appear in some positive literal (a literal not preceded by \code{not}) in the body.
If a variable appears positively in some predicate, then we say that this predicate binds the variable.

@@

\subsubsection{Classical Negation}\label{subsec:gringo:negation}

@@

\subsubsection{Disjunction}\label{subsec:gringo:disjunction}

Disjunctive logic programs permit connective~``\code{|}'' between atoms in rule heads.
A disjunction is true if at least one of its atoms is true.
Additionally, logic programs have to satisfy a minimality criterion, 
which we do not detail in this guide.
The simple program \code{a | b.} has the two answer sets $\{a\}$ and $\{b\}$ but
does not admit the answer set ${a,b}$ because it is no minimal model.

@@

\subsubsection{Built-In Arithmetic Functions}\label{subsec:gringo:arith}

@@

\subsubsection{Built-In Comparison Predicates}\label{subsec:gringo:comp}

@@

\subsubsection{Assignments}\label{subsec:gringo:assign}

@@

\subsubsection{Intervals}\label{subsec:gringo:interval}

@@

\subsubsection{Conditions}\label{subsec:gringo:condition}

@@

\subsubsection{Pooling}\label{subsec:gringo:pool}

@@

\subsubsection{Aggregates}\label{subsec:gringo:aggregate}

@@

An aggregate has a lower bound $l$, an upper bound $u$, an operation $\mathsf{op}$, 
and a multiset of literal $L_i$ each assigned to a weight $w_i$.
An aggregate is true if operation $\mathsf{op}$ applied to the multiset of weights 
of true literals is between the bounds (inclusive).
Currently, \gringo\ supports the aggregates 
\code{\#sum} (the sum of weights), 
\code{\#min} (the minimum weight), 
\code{\#max} (the maximum weight), and
\code{\#avg} (the average of all weights
\footnote{The average aggregate over an empty set of weights is defined to be always true irrespective of any bounds.}%
).
Furthermore, there are three aggregates that are syntactically different.
The first is the \code{\#count} aggregate:
\[l\,\code{\#count}\,\code{\char`\{}\, L_1\code{,} \dots\code{,} L_n \,\code{\char`\}}\, u\]%
which basically are \code{\#sum} aggregates with all weights set to one 
and duplicate true literals counted only once.
Finally, there are the two parity aggregates:
\begin{align*}
&\code{\#even}\,\code{\char`\{}\, L_1\code{,} \dots\code{,} L_n \,\code{\char`\}} & &
\code{\#odd}\,\code{\char`\{}\, L_1\code{,} \dots\code{,} L_n \,\code{\char`\}}
\end{align*}%
These aggregates are true if the number of different true literals is even or odd, respectively.

@@

\subsubsection{Optimization}\label{subsec:gringo:optimize}

Optimization statements extend the basic question of
whether a set of atoms is an answer set to
whether it is an optimal answer set.
To support this reasoning mode, \gringo\ and \clingo\ adopt the
optimization statements of \lparse~\cite{lparseManual},
indicated via keywords \const{\#maximize} and \const{\#minimize}.
As an optimization statement does not admit a body, any (local) variable
in it must also occur in an atom (over a domain or built-in predicate)
on the right-hand side of a condition (cf.\ Section~\ref{subsec:gringo:condition})
within the optimization statement.
In multiset notation (square brackets),
weights may be provided as with \const{\#sum} aggregates. %,
In set notation (curly brackets),
duplicates of literals are removed
as with \const{count} aggregates.
Additionally, priorities can be associated with each literal.
A (ground) optimize statement has the form:
\[\mathsf{opt}\,\code{[}\,L_1=w_1@p_1,\dots,L_n=w_n@p_n\,\code{\char`\}}\] 
\[\mathsf{opt}\,\code{\char`\{}\,L_1@p_1,\dots,L_n@p_n\,\code{\char`\}}\]
where $\mathsf{opt}$ is either \code{\#maximize} or \code{\#minimize} , $L_i$ are literals 
with associates (integer) weights $w_i$ and (integer) priorities $p_i$.

@@

\subsubsection{Meta-Statements}\label{subsec:gringo:meta}

@@

\paragraph{Comments.}
To keep records of the contents of a logic program,
a logic program file may include comments.
A comment until the end of a line is initiated by symbol ``\code{\%},''
and a comment within one or over multiple lines is enclosed
in ``\code{\%*}'' and ``\code{*\%}.''
As an abstract example, consider:
%
\begin{lstlisting}[numbers=none,escapechar=@]
@\textit{logic program}@  %* enclosed comment *%  @\textit{logic program}@
@\textit{logic program}@  % comment till end of line
@\textit{logic program}@
%*
comment over multiple lines
*%
@\textit{logic program}@
\end{lstlisting}

\paragraph{Hiding Predicates.}
Sometimes, one may be interested only in a subset of the atoms belonging
to an answer set.
In order to suppress the atoms of ``irrelevant'' predicates from the output,
the \code{\#hide} declarative can be used.
The meanings of the following statements are indicated via accompanying comments:
%
\begin{lstlisting}[numbers=none]
#hide.     % Suppress all atoms in output
#hide p/3. % Suppress all atoms of predicate p/3 in output
#hide p(X,Y) : q(X). % Supress p/3 if the condition holds
\end{lstlisting}
Note that for the conditionals on the right-hand side, 
the same conditions as described in Section \ref{subsec:gringo:condition} apply.

@@

\paragraph{Constant Replacement.}
Constants appearing in a logic program may actually be placeholders for
concrete values to be provided by a user.
An example of this is given in Section~\ref{subsec:ex:color}.
Via the \code{\#const} declarative,
one may define a default value to be inserted for a constant.
Such a default value can still be overridden via command line option
\code{--const} (cf.\ Section~\ref{subsec:opt:gringo}).
Syntactically, \code{\#const} must be followed by an assignment having
a (symbolic) constant on the left-hand side and a term on the right-hand side.
Some exemplary constant declarations are:
%
\begin{lstlisting}[numbers=none]
#const x = 42.
#const y = f(x,h).
\end{lstlisting}%
Note that (for efficiency reasons) constant declarations are order dependent.
In the example above \code{x} would be replaced by \code{42}
but when reversing the directives this would no longer be the case.

\paragraph{Domain Declarations.}
Usually, variable names are local to a rule,
where they must be bound via appropriate atoms.
This locality can be undermined by using
\code{\#domain} declarations 
that globally associate variable names to atoms.
An associated atom is then simply added to the body of a rule in which
such a predefined variable name occurs in.
The following is a made-up example:
%
\begin{lstlisting}[numbers=none]
p(1,1). p(1,2).
#domain p(X,Y).
#domain p(Y,Z).
q(Z,X) :- not p(Z,X).
\end{lstlisting}
%
The above program is a priori not safe because
variables~\var{X} and~\var{Z} are unbound in the last rule.
However, as they belong to \code{\#domain} declarations,
\gringo\ and \clingo\ expand the last rule to:
%
\begin{lstlisting}[numbers=none]
q(Z,X) :- p(X,Y), p(Y,Z), not p(Z,X).
\end{lstlisting}
%
Observe that the resulting program is safe.
Nevertheless, we suggest not to use domain statements because 
in ASP it is common to use very short variable names and
using domain statements likely results in name clashes
and undesired behavior.

\paragraph{Compute Statements.}
These statements are artifacts supported for backward compatibility.
Although we strongly recommend to avoid compute statements,
we now describe their syntax.
A compute statement is of the form
``$\const{\#compute}~n \code{\char`\{}\dots\code{\char`\}}\code{.}$''
(the non-negative integer $n$ is optional),
where the ``$\code{\char`\{}\dots\code{\char`\}}$'' part
is similar to a \const{\#count} aggregate.
The meaning is that all literals contained in
``$\code{\char`\{}\dots\code{\char`\}}$'' must hold w.r.t.\ answer sets
that are to be computed,
while~$n$ specifies a number of answer sets to compute.
As \clasp, \clingo, and \iclingo\ provide command line option
\code{--number} (cf.\ Section~\ref{subsec:opt:clasp})
to specify how many answer sets are to be computed,
they simply ignore~$n$.
Furthermore,
the ``$\code{\char`\{}\dots\code{\char`\}}$'' part can equivalently
be expressed in terms of integrity constraints,
as indicated in the comments provided along with the following example:
%
\begin{lstlisting}[xrightmargin=-20pt,numbers=none]
q(1;2).
{ p(1..5) }.
#compute 0 { p(X) : q(X) }.       % :- 1 { not p(X) : q(X) }.
#compute   { not p(X) : X=4..5 }. % :- 1 { p(X) : X=4..5 }.
\end{lstlisting}

\paragraph{External Statements.}
External statements are used to mark certain atoms as external. This means 
that those atoms are not subject of simplification and consequently are not removed 
from the logic program. There are two kinds of external directives, 
global and local external statements.

@@

\subsubsection{Integrated Scripting Language}\label{subsec:lang:lua}
Utilizing the scripting language Lua\footnote{\url{http://www.lua.org}}, 
\gringo's input language can be enriched by arbitrary arithmetical functions and implicit domains,
answer sets can be intercepted and for example inserted into a database, or
interactions between grounding and solving are possible when incrementally solving with \iclingo.
We do not give an introduction to Lua here (there are numerous tutorials on the web), 
but give some examples showing the capabilities of this integration.

\begin{example}
The first example shows basic Lua usage:
\lstinputlisting{examples/luaf.lp}
\marginlabel{%
To compute the unique answer set, invoke:\\
\code{%
\mbox{~}gringo -t \attach{examples/luaf.lp}{luaf.lp}}
}%
In Line 3 we add a function that calculates the greatest common divisor of two numbers.
This function is called in Line 20 and the result stored in predicate \code{gcd}/$3$.
Note that Lua function calls look like function symbols but are preceded by ``\code{@}''.
Regarding binding of variables, the same restrictions as with arithmetic in Section \ref{subsec:gringo:arith} apply.

In Line 9 we add a function that emulates a range term.
It returns a table containing all numbers in the interval $[\code{a},\code{b}]$.
The values in this table are then successively replaced for the call \code{\@rng(X,Y)}.
In fact, this function exactly behaves like a range term.
\eexample
\end{example}

\begin{table}[ht]
\centering
\begin{tabular}{||l|l||}
 \hline 
 \code{Val.NUM}  & Type identifier for \gringo\ numbers.\\
 \hline 
 \code{Val.ID}   & Type identifier for \gringo\ strings.\\
 \hline 
 \code{Val.FUNC} & Type identifier for function symbols.\\
 \hline 
 \code{Val.SUP}  & Type identifier for \gringo's \code{\#supremum}.\\
 \hline 
 \code{Val.INF}  & Type identifier for \gringo's \code{\#infimum}.\\
 \hline 
 \code{Val.new(type[,value][,args])}  & Creates new ground terms (see Example \ref{ex:lua:values}).\\
 \hline 
 \code{Val.cmp(a,b)}  & Compares two ground \gringo\ terms. \\
 \hline 
 \code{Val.type(a)} & Returns the type of a term. \\
 \hline 
 \code{Val.name(f)} & Returns the name of function symbol \code{f}. \\
 \hline 
 \code{Val.args(f)} & Returns the arguments of function symbol \code{f} \\
 \hline
\end{tabular}
\caption{The Val meta-table.\label{tab:val}}
\end{table}

@@

\begin{table}[ht]
\centering
\begin{tabular}{||l|p{6cm}||}
 \hline 
 \code{Assignment.begin(n,a)} & Starts iteration over an atom with name~\code{n} with arity~\code{a}. \\
 \hline 
 \code{Assignment.next()}     & Advances to the next atom in the assignment and returns false if there is none. \\
 \hline 
 \code{Assignment.args()}     & The arguments of the current atom. \\
 \hline 
 \code{Assignment.isTrue()}   & The current atom is true. \\
 \hline 
 \code{Assignment.isFalse()}  & The current atom is false. \\
 \hline 
 \code{Assignment.isUndef()}  & The current atom is undefined. \\
 \hline 
 \code{Assignment.level()}    & The decision level on which the current atom has been assigned. \\
 \hline
\end{tabular}
\caption{The Assignment meta-table.\label{tab:assignment}}
\end{table}

@@

\subsection{Input Language of \iclingo}\label{subsec:lang:iclingo}

System \iclingo~\cite{gekakaosscth08a} extends \clingo\ by an
\emph{incremental} computation mode that incorporates both grounding and solving.
Hence, its input language includes all constructs described
in Section~\ref{subsec:lang:gringo}.
In addition, \iclingo\ deals with statements of the following form:
%
\begin{lstlisting}[numbers=none,escapechar=@]
#base.
#cumulative @\textit{constant}@.
#volatile   @\textit{constant}@.
\end{lstlisting}
%
Via ``\code{\const{\#base}.},''
the subsequent part of a logic program is declared as static, that is,
it is processed only once at the beginning of an incremental computation.
In contrast, ``\code{\const{\#cumulative} \textit{constant}.}''
and ``\code{\const{\#volatile} \textit{constant}.}'' are used to declare
a (symbolic) \code{\textit{constant}} as a placeholder for incremental step numbers.
In the parts of a logic program below a \const{\#cumulative} statement,
\code{\textit{constant}} is in each step replaced with the current step number,
and the resulting rules, facts, and integrity constraints are accumulated
over a whole incremental computation.
While the replacement of \code{\textit{constant}} is similar,
a logic program part below a \const{\#volatile} statement is local to steps,
that is, all rules, facts, and integrity constraints computed in one step
are dismissed before the next incremental step.
Note that the type of a logic program part (static, cumulative, or volatile)
is determined by the last
\const{\#base}, \const{\#cumulative}, or \const{\#volatile} statement
preceding it.

>>>>

\subsubsection{Normal Programs and Integrity Constraints}\label{subsec:gringo:normal}
\index{Terms}
\index{Terms!Constants}
\index{Terms!Variables}
\index{Terms!Variables!Anonymous}
\index{Terms!Functions}
\index{Terms!\code{\#supremum}}
\index{Terms!\code{\#infimum}}

@@

\index{Statements!Rules}
\index{Statements!Facts}
\index{Statements!Integrity Constraints}
\begin{tabular}{rl@{}l}\label{eq:normal:rule}
\textbf{Rule:} & $A_0$& $\code{:-}~L_1\code{,}\dots\code{,}L_n$\code{.}
\\
\textbf{Fact:} & $A_0$&\code{.}
\\
\textbf{Integrity Constraint:} & & $\code{:-}~L_1\code{,}\dots\code{,}L_n$\code{.}
\end{tabular}
\\

\noindent
The head~$A_0$ of a fact or a rule is an \emph{atom} of the same form
as a function symbol or constant.
Any $L_j$ is a \emph{literal} of the form $A$ or $\code{not}~A$
for an atom~$A$ where the connective \code{not} corresponds to default negation\index{Literals!Default Negation}.
The set of literals $\{L_1,\dots,L_n\}$ is called the body of the rule.
Facts have an empty body.
Throughout this section we further extend the predicates that can be used 
in a rule including comparison predicates (Section \ref{subsec:gringo:comp}) and aggregates (Section \ref{subsec:gringo:aggregate}).
Furthermore, \gringo{} expects rules to be safe, i.e., 
all variables that appear in a rule have to appear in some positive literal (a literal not preceded by \code{not}) in the body.
If a variable appears positively in some predicate, then we say that this predicate binds the variable.

@@

\subsubsection{Classical Negation}\label{subsec:gringo:negation}
\index{Literals!Classical Negation}

@@

\subsubsection{Disjunction}\label{subsec:gringo:disjunction}
\index{Aggregates!Disjunction}
Disjunctive logic programs permit connective~``\code{|}'' between atoms in rule heads.
A disjunction is true if at least one of its atoms is true.
Additionally, logic programs have to satisfy a minimality criterion, 
which we do not detail in this guide.
The simple program \code{a | b.} has the two answer sets $\{a\}$ and $\{b\}$ but
does not admit the answer set ${a,b}$ because it is no minimal model.

@@

\subsubsection{Built-In Arithmetic Functions}\label{subsec:gringo:arith}
\index{Literals!Arithmetic Functions}
\index{Literals!Arithmetic Functions!Addition, \code{+}}
\index{Literals!Arithmetic Functions!Subtraction, \code{-}}
\index{Literals!Arithmetic Functions!Multiplication, \code{*}}
\index{Literals!Arithmetic Functions!Division, \code{/}}
\index{Literals!Arithmetic Functions!Modulo, \code{\%}}
\index{Literals!Arithmetic Functions!Exponentation, \code{**}}
\index{Literals!Arithmetic Functions!Absolute Value, \code{\#abs}}
\index{Literals!Arithmetic Functions!Bitwise AND, \code{\&}}
\index{Literals!Arithmetic Functions!Bitwise OR, \code{?}}
\index{Literals!Arithmetic Functions!Bitwise XOR, \code{\^}}
\index{Literals!Arithmetic Functions!Bitwise Complement, \code{\textasciitilde}}

@@

\subsubsection{Built-In Comparison Predicates}\label{subsec:gringo:comp}
\index{Literals!Comparison Predicates}
\index{Literals!Comparison Predicates!Equality, \code{==}}
\index{Literals!Comparison Predicates!Inequality, \code{!=}}
\index{Literals!Comparison Predicates!Less, \code{<}}
\index{Literals!Comparison Predicates!Less or Equal, \code{<=}}
\index{Literals!Comparison Predicates!Greater,\code{>}}
\index{Literals!Comparison Predicates!Greater or Equal\code{>=}}

@@

\subsubsection{Assignments}\label{subsec:gringo:assign}
\index{Literals!Assignments}
\index{Literals!Assignments!Term Unification, \code{:=}}
\index{Literals!Assignments!Variable Assignment, \code{=}}

@@

\subsubsection{Intervals}\label{subsec:gringo:interval}
\index{Terms!Intervals}

@@

\subsubsection{Conditions}\label{subsec:gringo:condition}
\index{Aggregates!Conditions}

@@

\subsubsection{Pooling}\label{subsec:gringo:pool}
\index{Terms!Pooling}

@@

\subsubsection{Aggregates}\label{subsec:gringo:aggregate}
\index{Aggregates}

@@

An aggregate has a lower bound $l$, an upper bound $u$, an operation $\mathsf{op}$, 
and a multiset of literal $L_i$ each assigned to a weight $w_i$.
An aggregate is true if operation $\mathsf{op}$ applied to the multiset of weights 
of true literals is between the bounds (inclusive).
Currently, \gringo\ supports the aggregates 
\code{\#sum} (the sum of weights), \index{Aggregates!Sum, \code{\#sum}}
\code{\#min} (the minimum weight), \index{Aggregates!Minimum, \code{\#min}}
\code{\#max} (the maximum weight), and \index{Aggregates!Maximum, \code{\#max}}
\code{\#avg} (the average of all weights \index{Aggregates!Average, \code{\#avg}}
\footnote{The average aggregate over an empty set of weights is defined to be always true irrespective of any bounds.}%
).
Furthermore, there are three aggregates that are syntactically different.
The first is the \code{\#count}\index{Aggregates!Count, \code{\#count}} aggregate:
\[l\,\code{\#count}\,\code{\char`\{}\, L_1\code{,} \dots\code{,} L_n \,\code{\char`\}}\, u\]%
which basically are \code{\#sum} aggregates with all weights set to one 
and duplicate true literals counted only once.
Finally, there are the two parity aggregates:
\index{Aggregates!Even Parity, \code{\#even}}%
\index{Aggregates!Odd Parity, \code{\#odd}}%
\begin{align*}
&\code{\#even}\,\code{\char`\{}\, L_1\code{,} \dots\code{,} L_n \,\code{\char`\}} & &
\code{\#odd}\,\code{\char`\{}\, L_1\code{,} \dots\code{,} L_n \,\code{\char`\}}
\end{align*}%
These aggregates are true if the number of different true literals is even or odd, respectively.

@@

\subsubsection{Optimization}\label{subsec:gringo:optimize}
\index{Statements!Optimize Statements}

Optimization statements extend the basic question of
whether a set of atoms is an answer set to
whether it is an optimal answer set.
To support this reasoning mode, \gringo\ and \clingo\ adopt the
optimization statements of \lparse~\cite{lparseManual},
indicated via keywords \const{\#maximize} and \const{\#minimize}.
As an optimization statement does not admit a body, any (local) variable
in it must also occur in an atom (over a domain or built-in predicate)
on the right-hand side of a condition (cf.\ Section~\ref{subsec:gringo:condition})
within the optimization statement.
In multiset notation (square brackets),
weights may be provided as with \const{\#sum} aggregates. %,
In set notation (curly brackets),
duplicates of literals are removed
as with \const{count} aggregates.
Additionally, priorities can be associated with each literal.
A (ground) optimize statement has the form: 
\index{Statements!Optimize Statements!\#minimize}%
\index{Statements!Optimize Statements!\#maximize}%
\[\mathsf{opt}\,\code{[}\,L_1=w_1@p_1,\dots,L_n=w_n@p_n\,\code{\char`\}}\] 
\[\mathsf{opt}\,\code{\char`\{}\,L_1@p_1,\dots,L_n@p_n\,\code{\char`\}}\]
where $\mathsf{opt}$ is either \code{\#maximize} or \code{\#minimize} , $L_i$ are literals 
with associates (integer) weights $w_i$ and (integer) priorities $p_i$.

@@

\subsubsection{Meta-Statements}\label{subsec:gringo:meta}
\index{Meta-Statements}

@@

\paragraph{Comments.}
\index{Meta-Statements!Comments}
To keep records of the contents of a logic program,
a logic program file may include comments.
A comment until the end of a line is initiated by symbol ``\code{\%},''
and a comment within one or over multiple lines is enclosed
in ``\code{\%*}'' and ``\code{*\%}.''
As an abstract example, consider:
%
\begin{lstlisting}[numbers=none,escapechar=@]
@\textit{logic program}@  %* enclosed comment *%  @\textit{logic program}@
@\textit{logic program}@  % comment till end of line
@\textit{logic program}@
%*
comment over multiple lines
*%
@\textit{logic program}@
\end{lstlisting}

\paragraph{Hiding Predicates.}
\index{Meta-Statements!Hiding Predicates, \code{\#show}, \code{\#hide}}%
Sometimes, one may be interested only in a subset of the atoms belonging
to an answer set.
In order to suppress the atoms of ``irrelevant'' predicates from the output,
the \code{\#hide} declarative can be used.
The meanings of the following statements are indicated via accompanying comments:
%
\begin{lstlisting}[numbers=none]
#hide.     % Suppress all atoms in output
#hide p/3. % Suppress all atoms of predicate p/3 in output
#hide p(X,Y) : q(X). % Supress p/3 if the condition holds
\end{lstlisting}
Note that for the conditionals on the right-hand side, 
the same conditions as described in Section \ref{subsec:gringo:condition} apply.

@@

\paragraph{Constant Replacement.}
\index{Meta-Statements!Constant Replacement, \code{\#const}}%
Constants appearing in a logic program may actually be placeholders for
concrete values to be provided by a user.
An example of this is given in Section~\ref{subsec:ex:color}.
Via the \code{\#const} declarative,
one may define a default value to be inserted for a constant.
Such a default value can still be overridden via command line option
\code{--const} (cf.\ Section~\ref{subsec:opt:gringo}).
Syntactically, \code{\#const} must be followed by an assignment having
a (symbolic) constant on the left-hand side and a term on the right-hand side.
Some exemplary constant declarations are:
%
\begin{lstlisting}[numbers=none]
#const x = 42.
#const y = f(x,h).
\end{lstlisting}%
Note that (for efficiency reasons) constant declarations are order dependent.
In the example above \code{x} would be replaced by \code{42}
but when reversing the directives this would no longer be the case.

\paragraph{Domain Declarations.}
\index{Meta-Statements!Domain Declarations, \code{\#domain}}%
Usually, variable names are local to a rule,
where they must be bound via appropriate atoms.
This locality can be undermined by using
\code{\#domain} declarations 
that globally associate variable names to atoms.
An associated atom is then simply added to the body of a rule in which
such a predefined variable name occurs in.
The following is a made-up example:
%
\begin{lstlisting}[numbers=none]
p(1,1). p(1,2).
#domain p(X,Y).
#domain p(Y,Z).
q(Z,X) :- not p(Z,X).
\end{lstlisting}
%
The above program is a priori not safe because
variables~\var{X} and~\var{Z} are unbound in the last rule.
However, as they belong to \code{\#domain} declarations,
\gringo\ and \clingo\ expand the last rule to:
%
\begin{lstlisting}[numbers=none]
q(Z,X) :- p(X,Y), p(Y,Z), not p(Z,X).
\end{lstlisting}
%
Observe that the resulting program is safe.
Nevertheless, we suggest not to use domain statements because 
in ASP it is common to use very short variable names and
using domain statements likely results in name clashes
and undesired behavior.

\paragraph{Compute Statements.}
\index{Meta-Statements!Compute Statements, \code{\#compute}}%
These statements are artifacts supported for backward compatibility.
Although we strongly recommend to avoid compute statements,
we now describe their syntax.
A compute statement is of the form
``$\const{\#compute}~n \code{\char`\{}\dots\code{\char`\}}\code{.}$''
(the non-negative integer $n$ is optional),
where the ``$\code{\char`\{}\dots\code{\char`\}}$'' part
is similar to a \const{\#count} aggregate.
The meaning is that all literals contained in
``$\code{\char`\{}\dots\code{\char`\}}$'' must hold w.r.t.\ answer sets
that are to be computed,
while~$n$ specifies a number of answer sets to compute.
As \clasp, \clingo, and \iclingo\ provide command line option
\code{--number} (cf.\ Section~\ref{subsec:opt:clasp})
to specify how many answer sets are to be computed,
they simply ignore~$n$.
Furthermore,
the ``$\code{\char`\{}\dots\code{\char`\}}$'' part can equivalently
be expressed in terms of integrity constraints,
as indicated in the comments provided along with the following example:
%
\begin{lstlisting}[xrightmargin=-20pt,numbers=none]
q(1;2).
{ p(1..5) }.
#compute 0 { p(X) : q(X) }.       % :- 1 { not p(X) : q(X) }.
#compute   { not p(X) : X=4..5 }. % :- 1 { p(X) : X=4..5 }.
\end{lstlisting}

\paragraph{External Statements.}
\index{Meta-Statements!External Statements, \code{\#external}}%
External statements are used to mark certain atoms as external. This means 
that those atoms are not subject of simplification and consequently are not removed 
from the logic program. There are two kinds of external directives, 
global and local external statements.

@@

\subsubsection{Integrated Scripting Language}\label{subsec:lang:lua}
\index{Lua}%
Utilizing the scripting language Lua\footnote{\url{http://www.lua.org}}, 
\gringo's input language can be enriched by arbitrary arithmetical functions and implicit domains,
answer sets can be intercepted and for example inserted into a database, or
interactions between grounding and solving are possible when incrementally solving with \iclingo.
We do not give an introduction to Lua here (there are numerous tutorials on the web), 
but give some examples showing the capabilities of this integration.

\begin{example}
The first example shows basic Lua usage:
\lstinputlisting{examples/luaf.lp}
\marginlabel{%
To compute the unique answer set, invoke:\\
\code{%
\mbox{~}gringo -t \attach{examples/luaf.lp}{luaf.lp}}
}%
In Line 3 we add a function that calculates the greatest common divisor of two numbers.
This function is called in Line 20 and the result stored in predicate \code{gcd}/$3$.
Note that Lua function calls look like function symbols but are preceded by ``\code{@}''. 
\index{Terms!Lua Function Call}%
\index{Lua!Function Call}%
Regarding binding of variables, the same restrictions as with arithmetic in Section \ref{subsec:gringo:arith} apply.

In Line 9 we add a function that emulates a range term.
It returns a table containing all numbers in the interval $[\code{a},\code{b}]$.
\index{Lua!Term Insertion}%
The values in this table are then successively inserted for the call \code{\@rng(X,Y)}.
In fact, this function exactly behaves like a range term.
\eexample
\end{example}

\begin{table}[ht]
\centering
\index{Lua!Val Metatable}%
\begin{tabular}{||l|l||}
 \hline 
 \code{Val.NUM}  & Type identifier for \gringo\ numbers.\\
 \hline 
 \code{Val.ID}   & Type identifier for \gringo\ strings.\\
 \hline 
 \code{Val.FUNC} & Type identifier for function symbols.\\
 \hline 
 \code{Val.SUP}  & Type identifier for \gringo's \code{\#supremum}.\\
 \hline 
 \code{Val.INF}  & Type identifier for \gringo's \code{\#infimum}.\\
 \hline 
 \code{Val.new(type[,value][,args])}  & Creates new ground terms (see Example \ref{ex:lua:values}).\\
 \hline 
 \code{Val.cmp(a,b)}  & Compares two ground \gringo\ terms. \\
 \hline 
 \code{Val.type(a)} & Returns the type of a term. \\
 \hline 
 \code{Val.name(f)} & Returns the name of function symbol \code{f}. \\
 \hline 
 \code{Val.args(f)} & Returns the arguments of function symbol \code{f} \\
 \hline
\end{tabular}
\caption{The Val meta-table.\label{tab:val}}
\end{table}

@@

\begin{table}[ht]
\index{Lua!Assignment Metatable}%
\centering
\begin{tabular}{||l|p{6cm}||}
 \hline 
 \code{Assignment.begin(n,a)} & Starts iteration over an atom with name~\code{n} with arity~\code{a}. \\
 \hline 
 \code{Assignment.next()}     & Advances to the next atom in the assignment and returns false if there is none. \\
 \hline 
 \code{Assignment.args()}     & The arguments of the current atom. \\
 \hline 
 \code{Assignment.isTrue()}   & The current atom is true. \\
 \hline 
 \code{Assignment.isFalse()}  & The current atom is false. \\
 \hline 
 \code{Assignment.isUndef()}  & The current atom is undefined. \\
 \hline 
 \code{Assignment.level()}    & The decision level on which the current atom has been assigned. \\
 \hline
\end{tabular}
\caption{The Assignment meta-table.\label{tab:assignment}}
\end{table}

@@

\subsection{Input Language of \iclingo}\label{subsec:lang:iclingo}
\index{Incremental Grounding}%

System \iclingo~\cite{gekakaosscth08a} extends \clingo\ by an
\emph{incremental} computation mode that incorporates both grounding and solving.
Hence, its input language includes all constructs described
in Section~\ref{subsec:lang:gringo}.
In addition, \iclingo\ deals with statements of the following form:
%
\begin{lstlisting}[numbers=none,escapechar=@]
#base.
#cumulative @\textit{constant}@.
#volatile   @\textit{constant}@.
\end{lstlisting}
%
\index{Incremental Grounding!Base Part, \code{\#base}}%
\index{Meta-Statements!Base Part, \code{\#base}}%
Via ``\code{\const{\#base}.},''
the subsequent part of a logic program is declared as static, that is,
it is processed only once at the beginning of an incremental computation.
In contrast, ``\code{\const{\#cumulative} \textit{constant}.}''
and ``\code{\const{\#volatile} \textit{constant}.}'' are used to declare
a (symbolic) \code{\textit{constant}} as a placeholder for incremental step numbers.
In the parts of a logic program below a \const{\#cumulative} statement,
\code{\textit{constant}} is in each step replaced with the current step number,
and the resulting rules, facts, and integrity constraints are accumulated
over a whole incremental computation.
While the replacement of \code{\textit{constant}} is similar,
a logic program part below a \const{\#volatile} statement is local to steps,
that is, all rules, facts, and integrity constraints computed in one step
are dismissed before the next incremental step.
Note that the type of a logic program part (static, cumulative, or volatile)
is determined by the last
\index{Incremental Grounding!Cumulative Part, \code{\#cumulative}}%
\index{Meta-Statements!Cumulative Part, \code{\#cumulative}}%
\index{Incremental Grounding!Volatile Part, \code{\#volatile}}%
\index{Meta-Statements!Volatile Part, \code{\#volatile}}%
\const{\#base}, \const{\#cumulative}, or \const{\#volatile} statement
preceding it.
